---
title: "Final Class Day: Data Quickfire!"
author: "Dr. Cassy Dorff"
date: "12/3/2020"
output:
  html_document:
    keep_md: true
urlcolor: blue
subtitle: DSI-Explore
editor_options: 
  chunk_output_type: console
---

Today you will compete in a DATA QUICKFIRE! What is at stake?

The winner:
 - One fashionable data science institute sweatshirt
 - One bonus point added to your final project's score
 
The runner up: 
  - Glory!FAME!
  - One bonus point added ot your final project's score

The goal: Create a *very* mis-leading graphic. Include a couple of sentences as to why it is misleading. 

Data options:

- covdata package (see below)
- john hopkins data (see below)
- covid.csv (from week 13)

## Covdata package

```{r}
# recommended
library(janitor)
library(stats)
library(tidyverse)

# data you need
install.packages("covdata")
library(covdata)
```

Where I found out about this data: https://github.com/mine-cetinkaya-rundel/covid19-r

How I installed it: 
- devtools::install_github("https://github.com/kjhealy/covdata", quiet=TRUE)
- documentation: https://github.com/kjhealy/covdata
- vignettes from package creator: https://github.com/kjhealy/covdata/tree/main/vignettes

Some example code below from the package maintaner:

```{r}
# data sets
head(covnat)
head(apple_mobility)
head(covus)

# Example from package maintainer
nytcovcounty %>%
  mutate(uniq_name = paste(county, state)) %>% # Can't use FIPS because of how the NYT bundled cities
  group_by(uniq_name) %>%
  mutate(days_elapsed = date - min(date)) %>%
  ggplot(aes(x = days_elapsed, y = cases, group = uniq_name)) + 
  geom_line(size = 0.25, color = "gray20") + 
  scale_y_log10(labels = scales::label_number_si()) + 
  guides(color = FALSE) + 
  facet_wrap(~ state, ncol = 5) + 
  labs(title = "COVID-19 Cumulative Recorded Cases by US County",
       subtitle = paste("New York is bundled into a single area in this data.\nData as of", format(max(nytcovcounty$date), "%A, %B %e, %Y")),
       x = "Days since first case", y = "Count of Cases (log 10 scale)", 
       caption = "Data: The New York Times | Graph: @kjhealy") + 
  theme_minimal()
```

## Johns Hopkins Data

The data is on github: https://github.com/CSSEGISandData/COVID-19

I have pulled data for you for the US below. You will need to do a bit of cleaning before working with it. There is actually example code of this on their github, I don't care if you use it, but if it slows you down, consider using a different data option.

```{r}
# IMPORT RAW DATA: Johns Hopkins Github data

jh_data <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")

# DATA CLEANING: To create country level and global combined data
# you would need to convert each data set from wide to long AND aggregate at country level
```

## Homework covid data

You may also choose to use the hw data from our clustering hw. Up to you!

## When done:

Submit on google drive
https://docs.google.com/document/d/10AethATm99UQpdNF1gC4aSKtQiGjkPhDDwU20UD8IoY/edit?usp=sharing


